Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2018-12-03 article_springer/article_springer.Rmd
Model Choice little changes

The classical POT approach is based on the fact, that exceedances above a high threshhold are assymptotically GPD distributed, hence a GPD distribution is fitted to the exceedances above a high enogh threshhold. Due to the Theorem in the last Section we know that for high threshholds the times between the exceedances are assymptotically Mittag-Leffler distributed in case of heavy tailed waiting times between the observations. Hence, before we talk about inference for the exceedances in the next section, we want to have a look at inference for Mittag-Leffler distributions in the following.

The first method proposed for the estimation of the Mittag-Leffler distribution parameters was the fractional moment estimator by [@kozubowski2001]. Unlike the first moments, the fractional moments of order $p$ for $p<\beta$ exist and can get calculated. One drawback of this method is that constant priors for the tail parameter are needed for the calculation of the estimates. [@Cahoy2010] proposed an estimator based on the log-Moments which doesn't need prior knowledge. Furthermore, they showed in a simulation study that the log-Moment outclasses the fractional moment estimator regarding bias and root mean squared error (RMSE).  
Due to the form of the density function of a Mittag-Leffler distribution there exists no closed form for the Maximum Likelihood estimator (MLE). In the RPackage `MittagLeffleR` Maximum Likelihood estimation is implemented via numerical optimization. The MLE outperforms the log-moment estimator regarding bias and RMSE for big enough sample sizes, but is extremely computational intensive and the differences in bias and RMSE between the two methods are very small. In Figure \ref{fig:MSE}, one can see that both methods work quite well even for very low sample sizes.    
```{r MSE, echo=FALSE, fig.height=12, fig.cap="\\label{fig:MSE} MSE for sample sizes n=30, 100, 200 for the tail and scale estimations via log-moments and maximum likelihood estimation", out.width = '100%'}
knitr::include_graphics("MSE_n25_100_200.pdf")
```
Since the Mittag-Leffler distribution is heavy-tailed, many researchers would intuitively give the highest importance to the tail behaviour of the distribution. Of course, one can also use established tail exponent estimators for the estimation of the parameter $\beta$ such as the Hill estimator, which is defined by 
\begin{align*}
H_{k,n}= \frac{1}{k} \sum_{i=0}^{k-1} \log \frac{X_{n-i,n}}{X_{n-k,n}}. 
\end{align*}
However, these methods are less statistically efficient since they only use a portion of the information contained in the data, as was already mentioned by [@kozubowski2001]. Additionally, they need a specification of a tuning parameter and the hill estimator only performs well for distributions close to Pareto (see @Resnick97). In Figure \ref{fig:Hillplots} one can see different Hill plots for Mittag-Leffler simulated data with differrent sample sizes and tails. It would be really hard to deduce the correct tail parameter estimates from this plots.

@@

```{r likelihood-ratio-test}
flares_arrivals <- flares %>% ctre() %>% thin(k = 500) %>% interarrival()
mlmle_out <- mlmle(data = flares_arrivals)
```

@@

It is beyond the scope of the current paper to incorporate parameter 
uncertainty into our predictive distribution for the next threshold crossing; 
however, methods as described by @Scarrott12 and @Lee15 are likely 
to extend to our setting.

@@

To test our inference method via stability plots, we have simulated 
independent waiting time and magnitude pairs $(W_k, J_k)$ (see upper panel in 
Figure \ref{fig:thresholdedBursty}) for three different situations: stable waiting times, Pareto waitings times and Exponential waiting times. In order to have exact 
analytical values available for $\beta$ and $\sigma_0$, a distribution for 
$W_k$ needs to be chosen for which $b(n)$ from \eqref{eq:stability} is known. 
If we choose $W_k \stackrel{d}{=} D$, where $D$ is as in \eqref{eq:stability}, 
then due to the stability property we have the *equality* of distribution
$W_1 + \ldots + W_n \stackrel{d}{=} b(n) D$, 
for $b(n) = n^{1/\beta}$. 
Using the parametrisation of @SamorodnitskyTaqqu, a few lines of 
calculation (see e.g. the vignette on parametrisation in @MittagLeffleR) 
show that $D$ must have the stable distribution 
$S_\beta(\cos(\pi \beta/2)^{1/\beta}, +1, 0)$, which is 
implemented in the R package `stabledist` by @stabledist.

@@

The "complete Hard X Ray Burst Spectrometer event list" 
is a comprehensive reference for all measurements of the Hard X 
Ray Burst Spectrometer on NASA's Solar Maximum Mission from the time of 
launch on Feb 14, 1980 to the end of the mission in Dec 1989. 12,776 
events were detected, with the "vast 
majority being solar flares". The list includes the start time, peak time, 
duration, and peak rate of each event. We have used "start time" as the 
variable for event times, and "peak rate" as the variable for event 
magnitudes.

## Checking Model Assumptions

The CTRE model is based on three main assumptions, which are repeated 
below. For each assumption, we suggest one means of checking if it holds:

@@

Figures \ref{fig:flare-diagnostics-1} and \ref{fig:flare-diagnostics-2} 
show the diagnostic plots for a minimum threshold chosen at the 200th order statistic. 
There is some residual autocorrelation for the sequence of threshold exceedance times that is not accounted for by the CTRE model. The fit with a Mittag-Leffler distribution ($\beta = 0.8$) is good, though there are signs that the power-law tail tapers off for very large inter-threshold crossing times. There is no apparent dependence between threshold exceedance times and event magnitudes seen in the copula plot.

@@

```{r flare-diagnostics-2, fig.height=4.5, fig.width=8, fig.cap="Diagnostic plots for the solar flare data: empirical copula and QQ Plot. \\label{fig:flare-diagnostics-2}"}
par(mfrow = c(1,2))
empcopula(flares_ctre)
flares_ctre %>% interarrival() %>% mlqqplot(log = 'xy', tail = 0.8, main = "Mittag-Leffler QQ Plot")
```

@@

@@

When we introduced the CTRE model, we assumed that all events are i.i.d. 
This assumption is likely sufficient but not necessary for our limit theorem to hold. 
Moreover, any data below a (minimum) threshold $\ell_0$ is discarded for 
CTREs, and hence need not satisfy the i.i.d. assumption.
For the purposes of statistical inference, we merely require that the 
inter-threshold-crossing times are i.i.d.

@@

Finally, a purely scale-free pattern for event times may be too rigid an 
assumption for some bursty time series, 
because often the heavy-tailed character of the inter-arrival time
distribution does not hold at all time scales; rather, it applies at 
short and intermediate time scales, and is truncated (or tempered, 
reverting to an exponential distribution) at 
very long time scales [see e.g. @MeerschaertRoyQin; and @Aban06]. 
In such situations, a "tempered" Mittag-Leffler distribution may provide 
a more realistic fit, which we aim to introduce in follow-up work.

>>>>

The classical peaks over threshold approach is based on the fact, that exceedances above a high threshhold are assymptotically GPD distributed. Hence, a GPD distribution is fitted to the exceedances above a high enough threshhold. Due to the Theorem in the last Section, we know that for high threshholds the times between the exceedances are in case of heavy tailed inter-arrival times assymptotically Mittag-Leffler distributed. Therefore, before talking about inference for the exceedance times in the next section, we want to have a look at inference for Mittag-Leffler distributions in general.

The first method proposed in the literature to estimate the parameters of Mittag-Leffler distribution was the fractional moment estimator by [@kozubowski2001]. Unlike the first moments, the fractional moments of order $p$ for $p<\beta$ exist and can get calculated by explicit formulas. One drawback of this method is that constant priors for the tail parameter are needed for the calculation of the estimates. Therefore, [@Cahoy2010] proposed an estimator based on the log-Moments which doesn't need prior knowledge. Furthermore, they showed in a simulation study that the log-Moment outclasses the fractional moment estimator regarding bias and root mean squared error (RMSE).  
Due to the form of the density function of a Mittag-Leffler distribution there exists no closed form for the Maximum Likelihood estimator (MLE). In the RPackage `MittagLeffleR` we implemented Maximum Likelihood estimation via an numerical optimization algorithm. The MLE outperforms the log-moment estimator regarding bias and RMSE for big enough sample sizes, but is extremely computational intensive and the differences in bias and RMSE between the two methods are very small. In Figure \ref{fig:MSE}, one can see that both methods work quite well even for very low sample sizes.    
```{r MSE, echo=FALSE, fig.height=12, fig.cap="\\label{fig:MSE} MSE for sample sizes n=30, 100, 200 for the tail and scale estimations via log-moments and maximum likelihood estimation", out.width = '100%'}
knitr::include_graphics("MSE_n25_100_200.pdf")
```
Since the Mittag-Leffler distribution is heavy-tailed, many researchers would intuitively give the highest importance to the tail behaviour of the distribution. Of course, one can also use established tail exponent estimators for the estimation of the parameter $\beta$ such as the Hill estimator, which is defined by 
\begin{align*}
H_{k,n}= \frac{1}{k} \sum_{i=0}^{k-1} \log \frac{X_{n-i,n}}{X_{n-k,n}}. 
\end{align*}
However, these methods are less statistically efficient since they only use a portion of the information contained in the data, as was already mentioned by [@kozubowski2001]. Additionally, they need a specification of a tuning parameter and the hill estimator only performs well for distributions close to Pareto (see @Resnick97). In Figure \ref{fig:Hillplots} one can see different Hill plots for Mittag-Leffler simulated data with differrent sample sizes and tails. It would be really hard to deduce the correct tail parameter estimates from these plots.

@@

@@

@@

To test our inference method, we have simulated independent waiting time and magnitude pairs $(W_k, J_k)$. The distribution of $J_k$ is irrelevant for the inference on $\beta$ and $\sigma_0$ (it just plays a role for the rescaling of $\hat{\sigma}_0$, which gets done to receive an estimate for the scale parameter of the exceedances). Hence we choosed exponential distributed magnitudes in all considered examples. For the waitings times, we choosed a Stable, a Pareto and an Exponential distribution.   
In order to have exact analytical values available for $\beta$ and $\sigma_0$, a distribution for $W_k$ needs to be chosen for which $b(n)$ from \eqref{eq:stability} is known. In our first example, we choose $W_k \stackrel{d}{=} D$, where $D$ is as in \eqref{eq:stability}, then due to the stability property we have the *equality* of distribution $W_1 + \ldots + W_n \stackrel{d}{=} b(n) D$, for $b(n) = n^{1/\beta}$. Using the parametrisation of @SamorodnitskyTaqqu, a few lines of calculation (see e.g. the vignette on parametrisation in @MittagLeffleR) show that $D$ must have the stable distribution $S_\beta(\cos(\pi \beta/2)^{1/\beta}, +1, 0)$, which is implemented in the R package `stabledist` by @stabledist. If we choose $W_k$ to be Pareto distributed or more precisecly 
\begin{align*}
P(W_k>t)= C t^{-\beta}
\end{align*}
with $C:=(1/\Gamma(1-\beta))^{1/\beta}$ and $b(n)=n^{1/\beta}$ we get 
\begin{align*}
\frac{W_1 + \ldots + W_n}{b(n)} \stackrel{d}{\Rightarrow} D
\end{align*}
with $E(\exp(-sD))=\exp(-s^{\beta})$.

@@

We now want to apply the proposed method to a real data example, the solar flare data which was already mentioned in Section 1 and can be seen in Figure \ref{fig:thresholdedBursty}. The data were extracted from the "complete Hard X Ray Burst Spectrometer event list", a comprehensive reference for all measurements of the Hard X Ray Burst Spectrometer on NASA's Solar Maximum Mission from the time of launch on Feb 14, 1980 to the end of the mission in Dec 1989. 12,776 events were detected, with the "vast  majority being solar flares". The list includes the start time, peak time, duration, and peak rate of each event. We have used "start time" as the variable for event times, and "peak rate" as the variable for event magnitudes.

Before we apply the POT approach described in Section 5 to the solar flare data, we first have to check if all model assumptions are full-filled. The CTRE model is based on three main assumptions, which are repeated below. For each assumption, we suggest one means of checking if it holds:

@@

Figures \ref{fig:flare-diagnostics-1} and \ref{fig:flare-diagnostics-2} show the diagnostic plots for a minimum threshold chosen at the 200th order statistic. There is some residual autocorrelation for the sequence of threshold exceedance times that is not accounted for by the CTRE model.

@@

```{r flare-diagnostics-2, fig.height=4.5, fig.width=8, fig.cap="Diagnostic plots for the solar flare data: empirical copula and QQ Plot. \\label{fig:flare-diagnostics-2}"}
par(mfrow = c(1,2))
empcopula(flares_ctre)
flares_ctre %>% interarrival() %>% mlqqplot(log = 'xy', tail = 0.85, main = "Mittag-Leffler QQ Plot")
```

Figure \ref{fig:flares} shows the stability plots for the solar flare data, on the left for the tail parameter and on the right for the scale parameter. The stability plot for the tail stabilizes nicely round about 0.85 (dotted line), while it is a little bit harder to deduce an estimate for the scale parameter. Howerever, the stability plot for the scale parameter seems to stabilize round about $3 \times 10^7$.

@@

The fit with a Mittag-Leffler distribution ($\beta = 0.85$) is good, though there are signs that the power-law tail tapers off for very large inter-threshold crossing times. There is no apparent dependence between threshold exceedance times and event magnitudes seen in the copula plot. We also counduct a bootstrapped LRT for the null hypothesis of exponential distributed inter-arrival times and received a $p$-value of

```{r likelihood-ratio-test}
library("MittagLeffleR")
library("stats")

# ML_LRT_boot<-function(data,signi=0.05,m=1000,n){
# if(is.null(dim(data))){
# mle_data<-mlmle(data)                                                   #Calculate loglikelihoods for obserevd data
# teststat<-2*(mle_data$loglik-mle_data$null_loglik)}                     #Calculate teststatistic for obserevd data
# else{
# l<-apply(data,MARGIN=2,FUN=mlmle)                                       #Extension to multivariate datasets for simulation studies
# teststat<-2*(unlist(lapply(l,"[[",2))-unlist(lapply(l,"[[",3)))
# }
# 
# x<-replicate(m,rexp(n,rate=1))                                          #Sample data under Null
# l2<-apply(x,MARGIN=2,FUN=mlmle)                                         #Calculate loglikelihoods for each simulated data set
# test_stat_null<-2*(unlist(lapply(l2,"[[",2))-unlist(lapply(l2,"[[",3))) #Calculate teststatistic for each simulated data set
# q<-stats::quantile(test_stat_null,probs=1-signi)                               #Calculate 1-signi quantile of simulated distribution of the test statistic
# #decision<-as.numeric(teststat>q)                                       #Gives One if Hypothesis gets rejected, 0 otherwise
# f<-function(x,y){
#   r<-as.numeric(x>y)
#   return(r)
# }
# decision<-mapply(teststat,FUN=f,y=q)
# decision<-return(decision)
# }
# flares_arrivals <- flares %>% ctre() %>% thin(k = 700) %>% interarrival()
# p_value <- ML_LRT_boot(data=flares_arrivals,n=length(flares_arrivals))
```

@@

When we introduced the CTRE model, we assumed that all events are i.i.d. This assumption is likely sufficient but not necessary for our limit theorem to hold. Moreover, any data below a (minimum) threshold $\ell_0$ is discarded for CTREs, and hence need not satisfy the i.i.d. assumption.For the purposes of statistical inference, we merely require that the inter-threshold-crossing times are i.i.d.

@@

Finally, a purely scale-free pattern for event times may be too rigid as an 
assumption for some bursty time series, because often the heavy-tailed character of the inter-arrival time
distribution does not hold at all time scales; rather, it applies at short and intermediate time scales, and is truncated (or tempered, 
reverting to an exponential distribution) at very long time scales [see e.g. @MeerschaertRoyQin; and @Aban06]. In such situations, a "tempered" Mittag-Leffler distribution may provide a more realistic fit, which we aim to introduce in follow-up work.
